---
title: 'Explaining Risk Ratios'
author: 'Ryan Estrellado'
date: '2017-12-04'
slug: explaining_risk_ratios
categories: []
tags: []
draft: true 
---

- Begin developing a practical understanding of what risk ratio means 

Imagine that you're visiting three booths at the farmers market. Each booth carries some combination of apples, oranges, and pears, but you don't know exactly what each has on the day. If you magically new from history what the probability was of each farmer carrying apples, you could figure out how likely one farmer was to carry apples compared to the other two. If farmer John was three times more likely to carry apples than the other farmers, we'd say that farmer John's risk ratio for carrying apples was three. 

You probably already know more about risk ratios in a practical sense than you think. Since you understand the concept of percentages, you're already most of the way there. A risk ratio is a comparison of one group's percentage to another group's percentage. In the farmer's market example, we're comparing the percentage of farmer John's fruit that are apples to the percent of all the other farmer's fruit that are apples. Here's the example in the form a table:

```{r}
library(tidyverse)

farmers <- tibble(
  farmer = rep(c(rep("John", 3), rep("Jane", 3), rep("Lilly", 3))), 
  fruit = rep(c("apples", "oranges", "plums"), 3), 
  count = sample(c(25:100), 9)
) 

farmers
```

Remember that in this example we're comparing the liklihood (aka, the risk) of John having apples to everyone else having apples. So we'll first need to figure out what the liklihood is that John has apples. 

```{r}
john <- farmers %>% 
  filter(farmer == "John") %>% 
  mutate(percent_of_total = round(count / sum(count), 2))

john
```

So we now know that John has a 29 percent chance (or risk) of fruit. 